A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-19 tex/body.tex
more

Prior to constructing a model for $F\!$, the theory solver does not explicitly construct $\mu$-terms nor the mapping $\Val$.
Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
To recognize when the precondition of \rn{Acyclic} holds,
the procedure considers each equivalence class $\ec{t}$ containing a datatype constructor $\const{C}( t_1, \ldots, t_n )$.
It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all arguments of constructors in these equivalence classes recursively.
If while doing so it returns to $\ec{t}$, it deduces that \rn{Acyclic} applies. % to $t$.
To recognize when the precondition of \rn{Unique} holds,
the procedure considers the set $\SSS$ of all codatatype equivalence classes.
It simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $\SSS$,
while partitioning $\SSS$ into $\SSS_1, \ldots, \SSS_n$ based on the top-most symbol of constructor terms in these equivalence classes
and the equivalence of their arguments of type $\Nondata$.
It then partitions each set in this partition in the same manner recursively.
If %after doing so,
the resulting partition contains a set $\SSS_i$ containing distinct terms $u$
and $v$, it deduces that \rn{Unique} applies to $u$ and $v$.

@@

%\paragraph{Optimizations.}
The implementation of the decision procedure uses several optimizations which
follow the approach outlined by Barrett et al. \cite{barrett-et-al-2007}.
%and are not reflected in the presentation of the calculus.
We briefly mention the main ones.
%
Discriminators are part of the signature instead
of being abbreviations.
We have found that this leads to better performance because the alternative introduces terms more eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several
rules \cite{barrett-et-al-2007},
which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
If $\const s^k_{\negvthinspace j}( t ) \in \tEc$ and $t =
\const{C}_{\negvthinspace j}( u_1, \ldots, u_n )$, we immediately add
$\smash{\const s^k_{\negvthinspace j}( t ) \teq u_k}$ to $\Ec$, whereas the presented calculus
would apply \rn{Split} and \rn{Inject} before adding this equality. 
\begin{rep}  % lower priority to mention this
To reduce the number of unique constraints considered by the calculus, we
compute a normal form for constraints as a preprocessing step. In particular, we
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term
ordering, replace $\const{C}_{\negvthinspace j}( \vec t ) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$ by $t_k$, and replace
occurrences of discriminators $\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t ) )$ by $\top$ or $\bot$
based on whether $j = \jPrime$.
\end{rep}

>>>>

Prior to constructing a model for $F\!$, the theory solver does not explicitly construct $\mu$-terms nor the mapping $\Val$.
Instead, $\Val$ is computed implicitly by traversing the equivalence classes of $\Ec$ during phase~2.
To detect whether \rn{Acyclic} applies,
the procedure considers each equivalence class $\ec{t}$ containing a datatype constructor $\const{C}( t_1, \ldots, t_n )$.
It visits $\ec{t_1}, \ldots, \ec{t_n}$ and all arguments of constructors in these equivalence classes recursively.
If while doing so it returns to $\ec{t}$, it deduces that \rn{Acyclic} applies. % to $t$.
To recognize when the precondition of \rn{Unique} holds,
the procedure considers the set $\SSS$ of all codatatype equivalence classes.
It simultaneously visits the equivalence classes of arguments of constructor terms in each equivalence class in $\SSS$,
while partitioning $\SSS$ into $\SSS_1, \ldots, \SSS_n$ based on the top-most symbol of constructor terms in these equivalence classes
and the equivalence of their arguments of type $\Nondata$.
It then partitions each set in this partition in the same manner recursively.
If %after doing so,
the resulting partition contains a set $\SSS_i$ containing distinct terms $u$
and $v$, it deduces that \rn{Unique} applies to $u$ and $v$.

@@

%\paragraph{Optimizations.}
The implementation of the decision procedure uses several optimizations
following the lines of Barrett et al. \cite{barrett-et-al-2007}.
%and are not reflected in the presentation of the calculus.
We briefly mention the main ones.
%
Discriminators are part of the signature instead
of being abbreviations.
We have found that this leads to better performance because the alternative introduces terms more eagerly to $\tEc$.
Handling discriminators requires extending the decision procedure with several
rules \cite{barrett-et-al-2007},
which apply uniformly to both datatypes and codatatypes.
Also, selectors are collapsed eagerly:
If $\const s^k_{\negvthinspace j}( t ) \in \tEc$ and $t =
\const{C}_{\negvthinspace j}( u_1, \ldots, u_n )$, we immediately add
$\smash{\const s^k_{\negvthinspace j}( t ) \teq u_k}$ to $\Ec$, whereas the presented calculus
would apply \rn{Split} and \rn{Inject} before adding this equality. 
\begin{rep}  % lower priority to mention this
To reduce the number of unique constraints considered by the calculus, we
compute a normal form for constraints as a preprocessing step. In particular, we
replace $t \teq s$ by $s \teq t$ if $s$ is less than $t$ based on a term
ordering, replace $\const{C}_{\negvthinspace j}( \vec t ) \teq
\const{C}_{\negvthinspace\jPrime}( \vec u )$ with $\bot$ when $j \neq \jPrime$,
replace all selector terms of the form $\const{s}^{\,k}_{\negvthinspace j}(
\const{C}_{\negvthinspace j}(t_1,\ldots,t_n) )$ by $t_k$, and replace
occurrences of discriminators $\const{d}_{\negvthinspace j}(
\const{C}_{\negvthinspace\jPrime}( \vec t ) )$ by $\top$ or $\bot$
based on whether $j = \jPrime$.
\end{rep}
